F\# Data: \textnormal{Making structured data first-class citizens}

### 2015-11-17 pldi-2016/paper.tex
Fit everything in 11 pages

\begin{abstract}
Most modern applications interact with external services and access data in structured formats such 
as XML, JSON and CSV. Static type systems do not understand such formats making accessing data more 
cumbersome. Should we give up and leave the messy world of external data to dynamic typing and 
runtime checks? Of course, not!

We show how to integrate external data sources into the F\# type system. As most real-world data
do not come with an explicit schema, we develop a type inference algorithm that infers a type from 
representative sample documents and integrate it into the F\# type system using type providers.

@@

Applications for taking notes, searching train schedules or finding tomorrow's weather all 
communicate with one or more services over the network and present the aggregated data. 
Increasing number of such services provide REST-based end-points that return data as CSV, XML
or JSON. Despite numerous schematization efforts, most services do not come with an 
explicit schema. At best, the documentation provides sample responses for typical requests.

For example, OpenWeatherMap provides an end-point for getting current weather for a 
given city\footnote{See ``Current weather data'': \url{http://openweathermap.org/current}}. 
The documentation describes the URL parameters and shows one sample JSON to illustrate the typical
response structure. Using a standard library for working with JSON and HTTP, we might call the 
service and read the temperature as follows:
%
\begin{equation*}
\begin{array}{l}
 \kvd{let}~\ident{doc}=\ident{Http.Request}(\str{http://api.owm.org/?q=NYC}) \\
 \kvd{match}~\ident{JsonValue.Parse}(\ident{doc})~\kvd{with} \\
 |~\ident{Record}(\ident{root})\rightarrow \\
 \quad \kvd{match}~\ident{Map.find}~\str{main}~\ident{root}~\kvd{with} \\
 \quad |~\ident{Record}(\ident{main})\rightarrow \\
 \quad \quad \kvd{match}~\ident{Map.find}~\str{temp}~\ident{main}~\kvd{with} \\
 \quad \quad |~\ident{Number}(\ident{num})\rightarrow \ident{printfn}~\str{Nice \%f degrees!}~\ident{num} \\
 \quad \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\
 \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\
 |~\_\rightarrow \ident{failwith}~\str{Incorrect format} 
\end{array}
\end{equation*}
%
The code assumes that the response has a particular format described in the documentation. The
root node must be a record with a \str{main} field, which has to be another record containing
a numerical \str{temp} field. When the format is incorrect, the data access simply fails
with an exception.

Using the JSON type provider from F\# Data, we can write code with exactly the 
same functionality in two lines:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{W} = \ident{JsonProvider}\langl\str{http://api.owm.org/?q=NYC}\rangl \\[0.1em]
 \ident{printfn}~\str{Nice \%f degrees!}~(\ident{W.GetSample().Main.Temp})
\end{array}
\end{equation*}
%
$\ident{JsonProvider}\langl\str{...}\rangl$ invokes a type provider at 
compile-time with the URL as a sample. The type provider infers the structure of the response
and provides a type with a \ident{GetSample} method that returns a parsed JSON with nested
properties \ident{Main.Temp}, returning the temperature as a number.

@@

\item We describe a predictable type inference algorithm for structured data formats that
  underlies the type providers (\S\ref{sec:inference}) based on a \emph{common supertype}
  relation.

\item We give a formal model (\S\ref{sec:formal}) and use it to prove
  \emph{relativized type safety} for the type providers (\S\ref{sec:safety}).
  This adapts the ML-style type safety for the context of the web.
\end{itemize}

@@

We start with an informal overview that shows how F\# Data type providers simplify working with 
JSON, XML and CSV. We introduce the necessary aspects of F\# type providers along the 
way. The examples in this section illustrate a number of key properties of our type inference 
algorithm:

\begin{itemize}
\item Our mechanism is predictable. The user directly works with the inferred types and should 
  understand why a specific type was inferred from a given sample\footnote{In particular, we do 
  not use probabilistic methods where adding additional sample could change the shape of the type.}.

@@

The JSON format used in \S\ref{sec:introduction} is a populardata exchange format based on 
JavaScript data structures. The following is the definition of \ident{JsonValue} 
used earlier to represent JSON:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{JsonValue} = \\[0.1em]
 \quad|~ \ident{Number}~\kvd{of}~\ident{decimal} ~~|~ \ident{Null} \\[0.1em]
 \quad|~ \ident{String}~\kvd{of}~\ident{string} \qquad\, |~ \ident{Boolean}~\kvd{of}~\ident{bool} \\[0.1em]
 \quad|~ \ident{Record}~\kvd{of}~\ident{Map}\langl\ident{string}, \ident{JsonValue}\rangl \\[0.1em]
 \quad|~ \ident{Array}~\kvd{of}~\ident{JsonValue}[] \\[0.1em]
\end{array}
\end{equation*}
%
The earlier example used only a nested record containing a number. To demonstrate other 
aspects of the JSON type provider, we look at an example that also involves an array:
%
{\small{
\begin{verbatim}
  [ { "name":"Jan", "age":25 }, { "name":"Tomas" },
    { "name":"Alexander", "age":3.5 } ]
\end{verbatim}
}}
%
\noindent
Say we want to print the names with an age if it is available. The standard approach would 
be to pattern match on the parsed \ident{JsonValue}, check that the top-level node is a \ident{Array}
and iterate over the elements checking that each is a \ident{Record} with certain properties. We
would throw an exception or skip values in incorrect format. Even if we defined helper functions, 
we would still need to specify field names as strings, which is error prone and can not be 
statically checked.

Assuming \strf{people.json} is the above example and \ident{data} is a string containing
JSON in the same format, we can write:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{People}~=~\ident{JsonProvider}\langl\str{people.json}\rangl\hspace{1em} \\[0.6em]
 \kvd{for}~\ident{item}~\kvd{in}~\ident{People.Parse}(\ident{data})~\kvd{do}\\[0.1em]
 \quad\ident{printf}~\str{\%s }~\ident{item.Name}\\[0.1em]
 \quad\ident{Option.iter}~(\ident{printf}~\str{(\%f)})~\ident{item.Age}
\end{array}
\end{equation*}
%
The code achieves the same simplicity as when using dynamically typed languages, but it is statically 
type-checked. In contrast to the earlier example, we now use a local file \strf{people.json} as a 
sample for the type inference, but then processes data from another source.

\paragraph{Type providers.}
The notation $\ident{JsonProvider}\langl\str{people.json}\rangl$ passes a \emph{static parameter} 
to the type provider. Static parameters are resolved at compile-time and have to be constant. 
The provider analyzes the sample and generates a type that we name \ident{People}. In most F\# 
editors, the type provider is also executed at development-time and the provided types are 
used in auto-completion.

The \ident{JsonProvider} uses a type inference algorithm (\S\ref{sec:inference})  and
infers the following types from the sample:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Entity}~=  \\
 \quad \kvd{member}~\ident{Name}~:~\ident{string} \\
 \quad \kvd{member}~\ident{Age}~:~\ident{option}\langl \ident{decimal}\rangl \\[0.5em]
 \kvd{type}~\ident{People}~=  \\
 \quad \kvd{member}~\ident{GetSample}~:~\ident{unit}~\rightarrow~\ident{Entity}[] \\
 \quad \kvd{member}~\ident{Parse}~:~\ident{string}~\rightarrow~\ident{Entity}[] \\
\end{array}
\end{equation*}
%
The type \ident{Entity} represents the person. The field \ident{Name} is available for all
sample values and is inferred as \ident{string}. The field \ident{Age} is marked as optional,
because the value is missing in one sample. The two sample ages are an integer $25$ and a 
decimal $3.5$ and so the common inferred type is \ident{decimal}.

The type \ident{People} has two methods for reading data. \ident{GetSample} parses the
sample used for the inference and \ident{Parse} parses a JSON string containing data in the same 
format as the sample. Both methods return an array of \ident{Entity} values.

\paragraph{Error handling.}
In addition to the structure of the types, the type provider also specifies what code should be 
executed at run-time in place of \ident{item.Name} and other operations. The runtime behaviour is 
the same as in the earlier hand-written sample (\S\ref{sec:introduction}). A member access 
throws an exception if the document does not have the expected format.

Informally, the safety property (\S\ref{sec:safety}) states that if the inputs
are subtypes of some of the provided samples (i.e.~the samples are representative), then
no exceptions will occur. In other words, we cannot avoid all failures, but we can prevent some.
If the OpenWeatherMap changes the response format, the sample in \S\ref{sec:introduction}
will not re-compile and the user knows that the code needs to be changed. What this means for
the traditional ML type safety is discussed in \S\ref{sec:safety-discuss}.

\paragraph{The role of records.}
The sample code is easy to write thanks to the fact that most F\# editors provide code completion
when ``.'' is typed. The developer does not need to look at the sample JSON file to see what fields
are available for a person. To support this scenario, our inference algorithm prefers records 
(we also treat XML elements and CSV rows as records) and only infers types that require pattern 
matching as the last resort.

@@

Consider a simple extensible document format where a root element
{\ttfamily\small <doc>} can contain a number of document element, one of which is
{\ttfamily\small <heading>} representing headings:
%
{\small{
\begin{verbatim}
  <doc>
    <heading>Working with JSON</heading>
    <p>Type providers make this easy.</p>
    <heading>Working with XML</heading>
    <p>Processing XML is as easy as JSON.</p>
    <image source="xml.png" />
  </doc>
\end{verbatim}
}}
%
\noindent
The F\# Data library has been designed primarilly to simplify reading of data. For example,
say we want to print all headings in the document. The sample shows a part of the document structure 
(in particular the {\ttfamily\small <heading>} element), but it does not show all possible elements 
(say, {\ttfamily\small <table>}). Assuming the above document is \strf{sample.doc}, we can write:
%
\noindent
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Document}~=~\ident{XmlProvider}\langl\str{sample.doc}\rangl\hspace{1em} \\[0.5em]
 \kvd{let}~\ident{root}~=~\ident{Document.Load}(\str{http://.../another.doc})\\
 \kvd{for}~\ident{elem}~\kvd{in}~\ident{root.Doc}~\kvd{do}\\
 \quad\kvd{match}~\ident{elem.Heading}~\kvd{with} \\
 \quad|~\ident{Some text} \rightarrow \ident{printf}~\str{ - \%s }~\ident{text}\\
 \quad|~\ident{None} \rightarrow ()\\
\end{array}
\end{equation*}
%
The example iterates over a collection of elements \ident{elem} returned by \ident{root.Doc}. 
The type provides a typed access to elements known from the sample document and so we can
write \ident{elem.Heading}, which returns an optional value.

\paragraph{Open world and variant types.}
In the above example, we do not infer a union type representing a choice between 
heading, paragraph and image. By its nature, XML is extensible and the sample cannot include all 
possible nodes\footnote{Even when the document structure is defined using XML Schema, documents may 
contain elements prefixed with other namespaces.}. The type of \ident{elem} is thus a top type (we 
call it \emph{variant}) annotated with the statically known possibilities (heading, paragraph, image). 
It is mapped to the following F\# type:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Element}~=  \\[0.1em]
 \quad \kvd{member}~\ident{Heading}~:~\ident{option}\langl \ident{string} \rangl\\
 \quad \kvd{member}~\ident{Paragraph}~:~\ident{option}\langl \ident{string} \rangl\\
 \quad \kvd{member}~\ident{Image}~:~\ident{option}\langl \ident{Image} \rangl\\
\end{array}
\end{equation*}
%
The type provides a nice access to the elements known statically from the sample, but it encodes
the \emph{open-world assumption} -- the user also needs to handle the case when the value is none 
of the statically known elements.

As a consequence, the variant type can be seen as the top type. The above code will work
correctly on document that contains any elements, including those not listed in the sample. This
is extremely useful when reading data from external data sources, because the sample document only
needs to contain values relevant to the application. It is also worth noting that our type 
inference attempts to minimze the number of variant types that appear in the resulting type --
a variant is not used if there is another option (Corollary~\ref{thm:no-unions}).

@@

\subsection{Summary}
In addition to JSON and XML (\S\ref{sec:providers-json}, \S\ref{sec:providers-xml}), F\# Data also 
implements a type provider for CSV. We treat CSV data as lists of records (with field for each 
column) and so handling of CSV is just a special case of our inference algorithm.

Throughout the introduction, we used data sets that demonstrate typical issues that are frequent
in real-world inputs (missing data, inconsistent encoding of primitive values and heterogeneous 
structures). In our experience, these are the most common issues. The following JSON response with 
government debt information returned by the World Bank API\footnote{Available at 
\url{http://data.worldbank.org}} demonstrates all three problems:
%
{\small{
\begin{verbatim}
  [ { "page": 1, "pages": 5 },
    [ { "indicator": "GC.DOD.TOTL.GD.ZS",
        "date": "2012", "value": null },
      { "indicator": "GC.DOD.TOTL.GD.ZS",
        "date": "2010", "value": "35.14229" } ] ]
\end{verbatim}
}}
%
\noindent
First of all, the top-level element is a collection containing two values of different kind.
The first is a record with meta-data about the current page and the second is an array with data. 
The actual implementation supports a concept of heterogeneous collections (omitted in this paper
for space reasons) and provides a type with properties \ident{Record} for the former and 
\ident{Array} for the latter. Second, the \ident{value} is \kvd{null} for some records. Third, 
numbers can be represented in JSON as numeric literals (without quotes), but here, they are 
returned as string literals instead\footnote{This is often used to avoid non-standard numerical 
types of JavaScript.}.

@@

We distinguish between \emph{non-nullable types} that always have a valid value (written as 
$\hat{\sigma}$) and \emph{nullable types} that encompass missing and \kvd{null} values 
(written as $\sigma$). We write $\nu$ for record field names and $\nu_?$ for names that can be empty:

\noindent
\begin{equation*}
\begin{array}{rcl}
 \hat{\sigma} &\narrow{=}& \nu_?\; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \} \\[0.1em]
                &\narrow{|}& \ident{float} \lsep \ident{decimal} \lsep \ident{int} \lsep \ident{bool} \lsep \ident{string} 
 \\[0.6em] 
       \sigma &\narrow{=}& ~\hat{\sigma}~ \lsep \kvd{option}\langl \hat{\sigma} \rangl \lsep [\sigma] \\[0.1em]
              &\narrow{|}& \kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl \lsep ~\bot~ \lsep \kvd{null}
\end{array}
\end{equation*}
%
Non-nullable types include records (consisting of an optional name and fields with their types) 
and primitive types. Records arising from XML documents are named, while records used by 
JSON and CSV providers are unnamed.

@@

Any non-nullable type can be wrapped in the \kvd{option} constructor to explicitly permit the 
\kvd{null} value. Type providers map \kvd{option} to standard F\# option. A collection $[\sigma]$ 
is also nullable and missing values or \kvd{null} are treated as empty collections. The type 
$\kvd{null}$ is inhabited by the $\kvd{null}$ value (using an overloaded notation) and $\bot$ 
is the bottom type.

@@

\noindent
\begin{align}
  \label{eq:sub-prim}\tag{P1}
  \ident{float}\,:>\,\ident{decimal}\,&:>\,\ident{int}&\\[-0.2em]
  \label{eq:sub-null}\tag{P2}
  \sigma &:> \kvd{null}  &(\textnormal{iff}~\sigma \neq \hat{\sigma}) 
\end{align}

\noindent
\begin{align}
  \label{eq:sub-opt}\tag{P3}
  \kvd{option}\langl\hat{\sigma}\rangl &:> \hat{\sigma}  &(\textnormal{for all}~\hat{\sigma})\\[-0.2em]
  \label{eq:sub-opt-cov}\tag{P4}
  \kvd{option}\langl\hat{\sigma_1}\rangl &:> 
    \kvd{option}\langl\hat{\sigma_2}\rangl  &(\textnormal{if}~\hat{\sigma_1} :> \hat{\sigma_2}) \\[-0.2em]
  \label{eq:sub-col}\tag{P5}
  [\sigma_1] &:> [\sigma_2]  &(\textnormal{if}~\sigma_1 :> \sigma_2) \\[-0.2em]
  \label{eq:sub-bot}\tag{P6}
  \sigma &:> \bot  &(\textnormal{for all}~\sigma)\\[-0.2em]
  \label{eq:sub-var-top}\tag{P7}
  \kvd{any}\langl \sigma_1, \ldots, \sigma_n\rangl &:> \sigma 
\end{align}

>>>>

\begin{abstract}
Most modern applications interact with external services and access data in structured formats such 
as XML, JSON and CSV. Static type systems do not understand such formats, often making data access
more cumbersome. Should we give up and leave the messy world of external data to dynamic typing 
and runtime checks? Of course, not!

In this paper, we integrate external structured data into the F\# type system. As most real-world data
do not come with an explicit schema, we develop a type inference algorithm that infers a type from 
representative sample documents and integrate it into the F\# type system using type providers.

@@

Social network clients, applications for finding tomorrow's weather or searching train schedules
all communicate with one or more services over the network and display the resulting information. 
Increasing number of such services provide REST-based end-points that return data as CSV, XML
or JSON. Despite numerous schematization efforts, most services do not come with an 
explicit schema. At best, the documentation provides sample responses for typical requests.

For example, OpenWeatherMap provides an end-point for getting current weather for a 
given city\footnote{See ``Current weather data'': \url{http://openweathermap.org/current}}. 
The documentation shows one sample to illustrate the typical response. Using standard 
JSON and web libraries, we might write:
%
\begin{equation*}
\begin{array}{l}
 \kvd{let}~\ident{doc}=\ident{Http.Request}(\str{http://api.owm.org/?q=NYC}) \\
 \kvd{match}~\ident{JsonValue.Parse}(\ident{doc})~\kvd{with} \\
 |~\ident{Record}(\ident{root})\rightarrow \\
 \quad \kvd{match}~\ident{Map.find}~\str{main}~\ident{root}~\kvd{with} \\
 \quad |~\ident{Record}(\ident{main})\rightarrow \\
 \quad \quad \kvd{match}~\ident{Map.find}~\str{temp}~\ident{main}~\kvd{with} \\
 \quad \quad |~\ident{Number}(\ident{num})\rightarrow \ident{printfn}~\str{Lovely \%f!}~\ident{num} \\
 \quad \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\
 \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\
 |~\_\rightarrow \ident{failwith}~\str{Incorrect format} 
\end{array}
\end{equation*}
%
The code assumes that the response has a particular format described in the documentation. The
root node must be a record with a \str{main} field, which has to be another record containing
a numerical \str{temp} field. When the format is incorrect, the data access simply fails
with an exception.

Using the JSON type provider from F\# Data, we can write code with exactly the 
same functionality in two lines:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{W} = \ident{JsonProvider}\langl\str{http://api.owm.org/?q=NYC}\rangl \\[0.1em]
 \ident{printfn}~\str{Lovely \%f!}~(\ident{W.GetSample().Main.Temp})
\end{array}
\end{equation*}
%
$\ident{JsonProvider}\langl\str{...}\rangl$ invokes a type provider at 
compile-time with the URL as a sample. The type provider infers the structure of the response
and provides a type with a \ident{GetSample} method that returns a parsed JSON with nested
properties \ident{Main.Temp}, returning the temperature as a number.

@@

\item We describe a predictable type inference algorithm for structured data formats, 
  based on a \emph{common supertype} relation, that underlies the type providers 
  (\S\ref{sec:inference}).

\item We give a formal model (\S\ref{sec:formal}) and use it to prove
  \emph{relativized type safety} for the type providers (\S\ref{sec:safety}).
  This adapts the ML-style type safety for the context of the web.

\item The supplementary screencast show developer experience using 
  F\# Data with JSON, XML and CSV.
\end{itemize}

@@

We start with an informal overview that shows how F\# Data type providers simplify working with 
JSON and XML. We introduce the necessary aspects of F\# type providers along the 
way. The examples in this section illustrate a number of key properties of our type inference 
algorithm:

\begin{itemize}
\item Our mechanism is predictable. The user directly works with the inferred types and should 
  understand why a specific type was inferred from a given sample.\footnote{In particular, we do 
  not use probabilistic methods where adding an additional sample could completely change the 
  shape of the type.}

@@

The JSON format used in \S\ref{sec:introduction} is a popular data exchange format based on 
JavaScript data structures. The following is the definition of \ident{JsonValue} 
used earlier to represent JSON:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{JsonValue} = \\[0.1em]
 \quad|~ \ident{Number}~\kvd{of}~\ident{decimal} ~~~~|~ \ident{Boolean}~\kvd{of}~\ident{bool} \\[0.1em]
 \quad|~ \ident{String}~\kvd{of}~\ident{string} ~~\qquad\, |~ \ident{Null} \\[0.1em]
 \quad|~ \ident{Record}~\kvd{of}~\ident{Map}\langl\ident{string}, \ident{JsonValue}\rangl \\[0.1em]
 \quad|~ \ident{Array}~\kvd{of}~\ident{JsonValue}[] \\[0.1em]
\end{array}
\end{equation*}
%
The earlier example used only a nested record containing a number. To demonstrate other 
aspects of the JSON type provider, we look at an example that also involves an array:
%
{\small{
\begin{verbatim}
  [ { "name":"Jan", "age":25 }, { "name":"Tomas" },
    { "name":"Alexander", "age":3.5 } ]
\end{verbatim}
}}
%
\noindent
Say we want to print the names and an age if it is available. The standard approach would 
be to pattern match on the parsed \ident{JsonValue}, check that the top-level node is a \ident{Array}
and iterate over the elements checking that each element is a \ident{Record} with certain properties. 
We would throw an exception for values in an incorrect format. Even with helper functions, 
we would still need to specify field names as strings, which is error prone and can not be 
statically checked.

Assuming \strf{people.json} is the above example and \ident{data} is a string containing
JSON in the same format, we can write:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{People}~=~\ident{JsonProvider}\langl\str{people.json}\rangl\hspace{1em} \\[0.6em]
 \kvd{for}~\ident{item}~\kvd{in}~\ident{People.Parse}(\ident{data})~\kvd{do}\\[0.1em]
 \quad\ident{printf}~\str{\%s }~\ident{item.Name}\\[0.1em]
 \quad\ident{Option.iter}~(\ident{printf}~\str{(\%f)})~\ident{item.Age}
\end{array}
\end{equation*}
%
In contrast to the earlier example, we now use a local file \strf{people.json} as a sample for 
the type inference, but then processes data from another source. The code achieves the same 
simplicity as when using dynamically typed languages, but it is statically type-checked.

\paragraph{Type providers.}
The notation $\ident{JsonProvider}\langl\str{people.json}\rangl$ passes a \emph{static parameter} 
to the type provider. Static parameters are resolved at compile-time and have to be constant. 
The provider analyzes the sample and generates a type  \ident{People}. Most F\# editors also 
execute the type provider in the background at development-time and use the provided types 
in auto-completion.

The \ident{JsonProvider} uses a type inference algorithm (\S\ref{sec:inference})  and
infers the following types from the sample:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Entity}~=  \\
 \quad \kvd{member}~\ident{Name}~:~\ident{string} \\
 \quad \kvd{member}~\ident{Age}~:~\ident{option}\langl \ident{decimal}\rangl \\[0.5em]
 \kvd{type}~\ident{People}~=  \\
 \quad \kvd{member}~\ident{GetSample}~:~\ident{unit}~\rightarrow~\ident{Entity}[] \\
 \quad \kvd{member}~\ident{Parse}~:~\ident{string}~\rightarrow~\ident{Entity}[] \\
\end{array}
\end{equation*}
%
The type \ident{Entity} represents the person. The field \ident{Name} is available for all
sample values and is inferred as \ident{string}. The field \ident{Age} is marked as optional,
because the value is missing in one sample. The two age values are an integer $25$ and a 
decimal $3.5$ and so the common inferred type is \ident{decimal}.

The type \ident{People} has two methods for reading data. \ident{GetSample} parses the
sample used for the inference and \ident{Parse} parses a JSON string. This lets us read
data at runtime, provided that it is in the same format as the static sample.

\paragraph{Error handling.}
In addition to the structure of the types, the type provider also specifies what code should be 
executed at run-time in place of \ident{item.Name} and other operations. The runtime behaviour is 
the same as in the earlier hand-written sample (\S\ref{sec:introduction}) -- a member access 
throws an exception if the document does not have the expected format.

Informally, the safety property (\S\ref{sec:safety}) states that if the inputs
are subtypes of some of the provided samples (i.e.~the samples are representative), then
no exceptions will occur. In other words, we cannot avoid all failures, but we can prevent some.
Moreover, if the OpenWeatherMap changes the response format, the sample in \S\ref{sec:introduction}
will not re-compile and the user knows that the code needs to be changed.

\paragraph{The role of records.}
The sample code is easy to write thanks to the fact that most F\# editors provide code completion
when ``.'' is typed (see the supplementary screencast). The developer does not need to look at the 
sample JSON file to see what fields are available. To support this scenario, our inference algorithm 
prefers records (we also treat XML elements and CSV rows as records) and only infers types that 
require pattern matching as the last resort.

@@

Consider a simple extensible document format where a root element
{\ttfamily\small <doc>} can contain a number of document element, one of which is
{\ttfamily\small <heading>} representing headings:
%
{\small{
\begin{verbatim}
  <doc>
    <heading>Working with JSON</heading>
    <p>Type providers make this easy.</p>
    <heading>Working with XML</heading>
    <p>Processing XML is as easy as JSON.</p>
    <image source="xml.png" />
  </doc>
\end{verbatim}
}}
%
\noindent
The F\# Data library has been designed primarilly to simplify reading of data. For example,
say we want to print all headings in the document. The sample shows a part of the document structure 
(in particular the {\ttfamily\small <heading>} element), but it does not show all possible elements 
(say, {\ttfamily\small <table>}). Assuming the above document is \strf{sample.xml}, we can write:
%
\noindent
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Document}~=~\ident{XmlProvider}\langl\str{sample.xml}\rangl\hspace{1em} \\[0.5em]
 \kvd{let}~\ident{root}~=~\ident{Document.Load}(\str{c:/pldi/another.xml})\\
 \kvd{for}~\ident{elem}~\kvd{in}~\ident{root.Doc}~\kvd{do}\\
 \quad\ident{Option.iter}~(\ident{printf}~\str{ - \%s})~\ident{elem.Heading}\\
\end{array}
\end{equation*}
%
The example iterates over a collection of elements returned by \ident{root.Doc}. The type of element 
provides a typed access to elements known from the sample and so we can write \ident{elem.Heading}, 
which returns an optional string value.

\paragraph{Open world and variant types.}
We do not infer a union type representing a choice between heading, paragraph and image. By its 
nature, XML is extensible and the sample cannot include all possible nodes.\footnote{Even when the 
document structure is defined using XML Schema, documents may contain elements prefixed with other 
namespaces.} The type of \ident{elem} is thus a top type (we call it a \emph{variant}) annotated 
with the statically known possibilities. It is mapped to the following F\# type:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Element}~=  \\[0.1em]
 \quad \kvd{member}~\ident{Heading}~:~\ident{option}\langl \ident{string} \rangl\\
 \quad \kvd{member}~\ident{Paragraph}~:~\ident{option}\langl \ident{string} \rangl\\
 \quad \kvd{member}~\ident{Image}~:~\ident{option}\langl \ident{Image} \rangl\\
\end{array}
\end{equation*}
%
The type provides a nice access to the statically known elements, but it encodes the 
\emph{open-world assumption}. The user also needs to handle the case when the value is none 
of the statically known elements.

As a consequence, the variant type can be seen as the top type. The above code will work
correctly on document that contains any elements, including those not listed in the sample. This
is extremely useful when reading data from external sources, because the sample document only
needs to contain values relevant to the application. It is also worth noting that our type 
inference attempts to minimze the number of variant types that appear in the resulting type --
a variant is not used if there is another option (see Corollary~\ref{thm:no-unions} and 
\S\ref{sec:impl-hetero}).

@@

\subsection{Summary}
\label{sec:providers-sum}
In addition to type providers for JSON and XM, F\# Data also implements a type provider for CSV 
(\S\ref{sec:impl-parsing}). We treat CSV files as lists of records (with field for each column) 
and so CSV is handled directly by our inference algorithm.

Throughout the introduction, we used data sets that demonstrate typical issues that are frequent
in the real-world (missing data, inconsistent encoding of primitive values and heterogeneous 
structures). In our experience, these are the most common issues. The following JSON response with 
government debt information returned by the World Bank API\footnote{Available at 
\url{http://data.worldbank.org}} demonstrates all three problems:
%
{\small{
\begin{verbatim}
  [ { "page": 1, "pages": 5 },
    [ { "indicator": "GC.DOD.TOTL.GD.ZS",
        "date": "2012", "value": null },
      { "indicator": "GC.DOD.TOTL.GD.ZS",
        "date": "2010", "value": "35.14229" } ] ]
\end{verbatim}
}}
%
\noindent
First of all, the top-level element is a collection containing two values of different kind.
The first is a record with meta-data about the current page and the second is an array with data. 
The actual F\# Data implementation supports a concept of heterogeneous collections (briefly outlined
in \S\ref{sec:impl-hetero}) and provides a type with properties \ident{Record} for the former and 
\ident{Array} for the latter. Second, the \ident{value} field is \kvd{null} for some records. Third, 
numbers can be represented in JSON as numeric literals (without quotes), but here, they are 
returned as string literals instead.\footnote{This is often used to avoid non-standard numerical 
types of JavaScript.}

@@

We distinguish between \emph{non-nullable types} that always have a valid value (written as 
$\hat{\sigma}$) and \emph{nullable types} that encompass missing and \kvd{null} values 
(written as $\sigma$). We write $\nu$ for record field names and $\nu_?$ for names that can be empty:
%
\begin{equation*}
\begin{array}{rcl}
 \hat{\sigma} &\narrow{=}& \nu_?\; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \} \\[0.1em]
                &\narrow{|}& \ident{float} \lsep \ident{decimal} \lsep \ident{int} \lsep \ident{bool} \lsep \ident{string} 
 \\[0.6em] 
       \sigma &\narrow{=}& ~\hat{\sigma}~ \lsep \kvd{option}\langl \hat{\sigma} \rangl \lsep [\sigma] \\[0.1em]
              &\narrow{|}& \kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl \lsep ~\bot~ \lsep \kvd{null}
\end{array}
\end{equation*}

\noindent
Non-nullable types include records (consisting of an optional name and fields with their types) 
and primitive types. Records arising from XML documents are named, while records used by 
JSON and CSV providers are unnamed.

@@

Any non-nullable type can be wrapped in the \kvd{option} constructor to explicitly permit the 
\kvd{null} value. Type providers map \kvd{option} to standard F\# option. A collection $[\sigma]$ 
is also nullable and \kvd{null} values are treated as empty collections. The type 
$\kvd{null}$ is inhabited by the $\kvd{null}$ value (using an overloaded notation) and $\bot$ 
is the bottom type.

@@

\noindent
\begin{align}
  \label{eq:sub-prim}\tag{P1}
  \ident{float}\,:>\,\ident{decimal}\,&:>\,\ident{int}&\\[-0.2em]
  \label{eq:sub-null}\tag{P2}
  \sigma &:> \kvd{null}  &(\textnormal{iff}~\sigma \neq \hat{\sigma})  \\[-0.2em]
  \label{eq:sub-opt}\tag{P3}
  \kvd{option}\langl\hat{\sigma}\rangl &:> \hat{\sigma}  &(\textnormal{for all}~\hat{\sigma})\\[-0.2em]
  \label{eq:sub-opt-cov}\tag{P4}
  \kvd{option}\langl\hat{\sigma_1}\rangl &:> 
    \kvd{option}\langl\hat{\sigma_2}\rangl  &(\textnormal{if}~\hat{\sigma_1} :> \hat{\sigma_2})\\[-0.2em]
  \label{eq:sub-col}\tag{P5}
  [\sigma_1] &:> [\sigma_2]  &(\textnormal{if}~\sigma_1 :> \sigma_2) \\[-0.2em]
  \label{eq:sub-bot}\tag{P6}
  \sigma &:> \bot  &(\textnormal{for all}~\sigma)\\[-0.2em]
  \label{eq:sub-var-top}\tag{P7}
  \kvd{any}\langl \sigma_1, \ldots, \sigma_n\rangl &:> \sigma 
\end{align}

\noindent
\begin{align}
\label{eq:sub-record1}\tag{R1}
\begin{array}{l}
 \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} :> \\
 \nu_?~\{ \nu_1\!:\!\sigma_1', .., \nu_n\!:\!\sigma_n' \}
\end{array} \qquad ~~~~(\textnormal{if}~\sigma_i :> \sigma_i')
\end{align}
\vspace{-1em}
\begin{align}
\label{eq:sub-record2}\tag{R2}
\begin{array}{l}
 \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} :> \\
 \nu_?~\{ \nu_1\!:\!\sigma_1, .., .., \nu_m\!:\!\sigma_m \}
\end{array} \qquad (\textnormal{when}~m \geq n)
\end{align}
\vspace{-1em}
\begin{align}
\label{eq:sub-record3}\tag{R3}
\begin{array}{l}
 \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} :> \\
 \nu_?~\{ \nu_{\pi(1)}\!:\!\sigma_{\pi(1)}, .., \nu_{\pi(m)}\!:\!\sigma_{\pi(m)} \}
\end{array} ~~(\pi~\textnormal{perm.})
\end{align}
\vspace{-1em}
\begin{align}
\label{eq:sub-record4}\tag{R4}
\begin{array}{l}
 \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n, \nu_{n+1}\!:\!\kvd{null} \} :> \\
 \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \}
\end{array}
\end{align}
\end{definition}
